Formalizing Stack Safety as a Security Property

### 2023-01-31 submission5/ideas-v1.tex
Moved things around in the introduction and went back to v1 of key ideas. Also trimmed down v1 even further.

\sna{Might like a better name than object, because morally it ought to apply to arguments as well.}

\paragraph*{The Properties and their Hierarchy}

The properties form a hierarchy, shown in [TODO: diagram].
Within properties that are orthogonal
some are likely to take priority in terms of their
importance to security. Almost all stack-safety enforcement mechanisms will try to enforce \(WBCF\),
as control-flow attacks such as ROP are a major vulnerability that can be targeted via the stack.
Many also cover some degree of (caller) integrity. Caller confidentiality has been fairly neglected,
and callee-oriented properties even moreso.

@@

@@

This is a strong threat model, but it does omit some important aspects of
security enforcement: In particular, hardware and timing attacks are out of scope.

>>>>

@@

\paragraph*{Variants and Non-interference}

We model confidentiality by drawing on

@@

This is a strong threat model, but it does omit some important aspects of
security enforcement: in particular, hardware and timing attacks are out of scope.

\paragraph*{Tagged Architectures and Laziness}

In \cref{sec:testing}, we apply our properties to test the existing stack-safety
mechanism of Roessler and DeHon \cite{}, who present a collection of tag-based enforcement
mechanisms using PIPE \cite{???}, a tagged hardware reference monitor.
In monitors such as PIPE and STAR \cite{???}, \apt{double-check this} every value in memory
and registers is paired with a {\em metadata tag}.
A set of software-defined rules, termed a micro-policy, determines the tags for operation
results based on the tags of their inputs---or, if the input tags are in a dangerous
combination, it may halt the machine (a ``failstop.'')

We are particularly interested in their {\em Lazy Tagging and Clearing} micro-policy.
In their conventional micro-policies, they tag each word
in a stack frame with the (depth of) its owning function activation.
Roessler and DeHon find that it is inefficient, because tag setting
and clearing must be performed at every allocation, even on elements that are never accessed,
or that are initialized again and accessed long afterward. This is quite common, e.g. in code
that allocates large local arrays.

Instead they offer Lazy Tagging and Clearing, in which tags are not initialized at allocation, but only
when a function writes to memory. This means that writes cannot be prevented, which in turn means
that the existing line of stack safety specifications from Georges et al. will deem it unsafe.
Instead the micro-policy enforces that memory be read by the same function activation that wrote it.
The owner of the data will never see the changes,
so it is secure in the sense that the attacker is unsuccessful at changing program behavior
(except, perhaps, by causing a fault). We elaborate this notion of observational safety, and test whether the
micro-policy fulfills it.

\paragraph*{Randomized Property-based Testing}

[TODO: Brief discussion of PBT and why we use it, explain the lack of proofs.]

### 2023-01-31 submission5/intro-v2.tex
Moved things around in the introduction and went back to v1 of key ideas. Also trimmed down v1 even further.

\rb{A.K.A. motivation}

Computer programming % at any scale
relies on \emph{functions} (variously known as
subroutines, procedures, methods, etc.) as a foundational abstraction.
\apt{General comment: I think it's better to reserve italics for terms being defined, and avoid using them just for rhetorical emphasis.}
Functions are units of computation that
can \emph{call} one another to realize larger computations.
%
At a low level, a function call \apt{``activation''?} comprises its own data (private or shared), as
well as information about its \emph{caller} and \emph{callee} functions. The
runtime must keep track of the local contexts \apt{not defined} of all function activations
at any given time; the fundamental data structure charged with realizing
this abstraction is the \emph{(call) stack}. But at the level of the ABI
\sna{Do we need to unpack the acronym ABI?}, the specification of a well-behaved
stack is minimal. It is not concerned with such questions as whether a caller's
frame should be readable or writable to its callee, leaving those concerns
to be determined by a high-level language built atop the stack.

The correct execution---and the security!---of a program is contingent on all of
its functions properly implementing the intended abstraction by means of a careful use
of the call stack. Be it accidental or intentional, incorrect management of the
stack is at the root of many faults whose effects range from (rarely)
innocuous to (usually) catastrophic.\apt{Long-winded.}
%
From a security perspective, attacks on the stack (say, from a malicious
function that chooses to abuse its access to the contents of this data
structure) are as fundamental, varied and pervasive as the functional
abstraction itself
%

\apt{I think this is all well-known.}

%% \rb{\textbf{The call stack and its security.} Maybe a few opening words to set
%%   the stage and bring two things to the front: (1) functions and subroutines as
%%   fundamental building blocks of any kind of structured programming; and (2) the
%%   stack as the natural abstraction to manage these in a computer. And also maybe
%%   a little bit more about the width, the gravity and the \emph{foundational
%%   nature} of attacks on the stack, which the sentence below touches on
%%   summarily.
%% \rb{(+SNA) Connection to functions (and adversarial code?). Attacks on the stack
%%   are also attacks on this fundamental abstraction (functions, subroutines,
%%   procedures, methods, however you call them)}
%% }

Indeed, the call stack is a perennial target for low-level attacks, leading to
dire consequences, from leakage and corruption of private stack data
to control-flow hijacking.
%
The most recent release of the authoritative CWE Top 25
Most Dangerous Software Weaknesses
list\footnote{\url{https://cwe.mitre.org/top25/archive/2022/2022_cwe_top25.html}}
shows general classes of vulnerabilities closely related to the stack \rb{(though not exclusively)}
consistently at and near the top, e.g., \#1 (CWE-787, Out-of-bounds Write), \#5 (CWE-125, Out-of-bounds Read), \#7 (CWE-416, Use After Free) or \#11 (CWE-476, NULL Pointer Dereference).
%
Moreover, many other types of faults involve some kind of control or data flow
using data stored in the stack \rb{cf. Eternal war in memory, Google/Microsoft
  memory safety statistics, etc.}.
\sna{Trying to think how to make the case that these CWEs are about the stack, since they are
  primarily described as memory attacks.}\apt{Agree: the out-of-bounds ones are often on the stack, but the NULL pointer deref is not.
Also, this story is very well known, and doesn't need belaboring.}
%
To foil such attacks, a menagerie of
software and hardware protections have been proposed,
%
including stack canaries~\cite{Cowan+98},
bounds checking~\cite{NagarakatteZMZ09,NagarakatteZMZ10,DeviettiBMZ08},
split stacks~\cite{Kuznetsov+14},
shadow stacks~\cite{Dang+15,Shanbhogue+19},
capabilities~\cite{Woodruff+14,Chisnall+15,SkorstengaardLocal,SkorstengaardSTKJFP,Georges+21},
and hardware tagging~\cite{DBLP:conf/sp/RoesslerD18}---we discuss these in more
detail in \cref{XXX}.
  \ifaftersubmission\apt{Mostly from
  nick; there could be more}\bcp{Yes, going back to MIT days---we should
  include several more of these, if only to give readers the impression that
this is a well-studied mechanism (so formalizing its protections is
useful).}
\fi
%
\rb{Broad enough? Do we want to add more techniques?}
\sna{I think this is plenty unless there's a really major thing we're forgetting.}
%
Yet despite this wealth of potential solutions, the stack remains a central
security concern---why?
\apt{Mainly because preventing common stack problems generally causes a performance hit.
  And, to a (much) lesser extent, because C/C++ don't carry length information with
  arrays.  The sequencing here seems to suggest that we are going to \emph{solve} the
  problem of enforcing stack safety by defining it better---which I don't buy at all!}

\paragraph{Formal Properties}

Security mechanisms designed to prevent a certain attack can be brittle---successfully
eliminating the attack as it exists at the time of publication, but leaving room for attackers
to find new, similar attacks.
%
%% \rb{And so the cat-and-mouse game continues.}
%
To mitigate this risk, we claim that mechanisms should aim to enforce \emph{formal
properties} of safe behavior that can be proven or tested. Such properties become the
\emph{specification} against which enforcement can be validated; even enforcement mechanisms that do
not fulfill a property benefit from the ability to articulate \emph{why} they fail.

%
%% \rb{Good stuff! There is some overlap between this discussion and the opening
%%   motivation, maybe we could restructure things a bit so that those threads are
%%   unified?}
%

\subsection{Related Work}

\apt{This section was a nice summary of enforcement approaches, but I am not
  yet seeing why it justifies so much paper real estate, especially if we only 
  actually apply our property definitions to (lazy) micropolicies. And in any case,
  it makes for a huge delay before getting to our work and contributions.}

@@

\subsection{A Language-Based View of Stack Safety}

@@

Stack safety is challenging to specify in full generality, because the stack is
fundamentally an architecture-specific construct.
(Contrast Azevedo de Amorim et al.'s \cite{DBLP:conf/post/AmorimHP18} work on heap safety,
which is broadly consistent across architectures and language semantics, and therefore
amenable to a very general approach.)
The desired behavior of the stack varies based on ISA, calling convention,
and the features of whichever programming language originated the code.
%
For instance, the current state-of-the-art stack-safety specification from Georges et al.
\cite{Georges22:TempsDesCerises} supposes a machine with only straightforward call-and-return
control flow, and therefore defines {\it Well-bracketed Control Flow} (WBCF) under the assumption that
a callee should always return to its caller. But, in the presence of tail-call elimination, a
return from a callee to a distant ancestor is perfectly reasonable.

In this work, we propose an explicit specification, built from first principles
in the hopes that it will be widely applicable, and demonstrate that it supports
an enforcement mechanism that does not yet have one.
We use the formal tools of language-based
security~\cite{sabelfeld2003language}, decomposing stack safety into
the {\em integrity} and {\em confidentiality} of the caller’s local state
and the callee's behavior during the callee's execution, plus the control-flow protection
of {\em well-bracketed control flow}~\cite{SkorstengaardSTKJFP} (\(\wbcf\)).
\ifexceptions
We also extend WBCF to a setting with exceptions,
in which there may be multiple points at which a callee can ``return.''
\fi
%
We treat stack safety as multiple properties, rather than one multi-faceted property, because
not all enforcement mechanisms attempt to implement all aspects of stack safety. Some focus
exclusively on control flow, some enforce integrity but not confidentiality, and many
focus on the caller over the callee.
%
\rb{Return to this after extended discussion? Forward ref?}
%
This gives us a total of five core properties: \(\wbcf\),
caller integrity (\(\clri\)), caller confidentiality (\(\clrc\)),
callee integrity (\(\clei\)), and callee confidentiality (\(\clec\)), defined formally
in \cref{sec:props}.
%

We are not attempting to create a universal definition of stack safety for all systems.
Rather, we define the stack safety of a single, fairly realistic system, decomposed into
(1) a lightweight model of the system and its security-relevant features, from which we
derive the integrity, confidentiality, and control-flow requirements of each call and
return, (2) the application of that model to a particular target system, and (3)
the formal criteria for integrity, confidentiality, and control flow
to obey the requirements. This decomposition can be thought of as a step toward
a universal definition, in that the definitions in (3) capture fundamental
notions of security that can then be applied flexibly to other systems by varying
the model in (1) and (2).
%
\rb{FIXME: Contrast some universal definitions (or a parametric framework) with
  the concrete instance we study, again with its interest properly motivated.
  (Coming in key ideas?) No ``gold standard''?}
\apt{Yes, there are at least two different kinds of parameterization going on.
  One has to do with which properties we want to, or are able to, enforce; e.g.,
  for a language with tail calls or exceptions, we need a more complex definition of the WBCF property itself.
  Another has to do with how standard features, like argument passing, are implemented
  by particular hardware/ABIs -- which shouldn't show up in the property definition.
  We need to be clearer about this.}

\sna{Not sure what your FIXME means exactly. Can you elaborate?}

Our artifact contains Coq formalizations of our properties for illustrative purposes
as well as their testing implementation. There are no proofs---we use primarily Coq for the
QuickChick library and to ensure that the formalizations are unambiguous.
%
\rb{Will have another look after the full pass.}

>>>>

Computer programming relies on functions (subroutines, procedures, methods, etc.)
as a foundational abstraction. Functions are units of computation that
can call one another to realize larger computations.
%
At a low level, a function activation comprises its own data, as
well as information about its \emph{caller} function, to which it must return.
The \emph{(call) stack} is the fundamental data structure charged with realizing
this abstraction. From a security perspective, attacks on the stack are attacks
on the fundamental functional abstraction itself, and are therefore every bit as
pervasive.

Indeed, the call stack is a perennial target for low-level attacks, leading to
dire consequences, from leakage and corruption of private stack data
to control-flow hijacking.
%
The most recent release of the authoritative CWE Top 25
Most Dangerous Software Weaknesses
list\footnote{\url{https://cwe.mitre.org/top25/archive/2022/2022_cwe_top25.html}}
shows general classes of vulnerabilities closely related to the stack \rb{(though not exclusively)}
consistently at and near the top, e.g., \#1 (CWE-787, Out-of-bounds Write), \#5 (CWE-125, Out-of-bounds Read), \#7 (CWE-416, Use After Free) or \#11 (CWE-476, NULL Pointer Dereference).
%
Moreover, many other types of faults involve some kind of control or data flow
using data stored in the stack \rb{cf. Eternal war in memory, Google/Microsoft
  memory safety statistics, etc.}.
\sna{Trying to think how to make the case that these CWEs are about the stack, since they are
  primarily described as memory attacks.}\apt{Agree: the out-of-bounds ones are often on the stack, but the NULL pointer deref is not.
Also, this story is very well known, and doesn't need belaboring.}
%
To foil such attacks, a menagerie of
software and hardware protections have been proposed,
%
including stack canaries~\cite{Cowan+98},
bounds checking~\cite{NagarakatteZMZ09,NagarakatteZMZ10,DeviettiBMZ08},
split stacks~\cite{Kuznetsov+14},
shadow stacks~\cite{Dang+15,Shanbhogue+19},
capabilities~\cite{Woodruff+14,Chisnall+15,SkorstengaardLocal,SkorstengaardSTKJFP,Georges+21},
and hardware tagging~\cite{DBLP:conf/sp/RoesslerD18}---we discuss these in more
detail in \cref{XXX}.
  \ifaftersubmission\apt{Mostly from
  nick; there could be more}\bcp{Yes, going back to MIT days---we should
  include several more of these, if only to give readers the impression that
this is a well-studied mechanism (so formalizing its protections is
useful).}
\fi
%
\rb{Broad enough? Do we want to add more techniques?}
\sna{I think this is plenty unless there's a really major thing we're forgetting.}
%

\paragraph{Towards a Formal Property for Stack Safety}

Security mechanisms designed to prevent a certain attack can be brittle---successfully
eliminating the attack as it exists at the time of publication, but leaving room for attackers
to find new, similar attacks.

To mitigate this risk, mechanisms should aim to enforce \emph{formal properties}
of safe behavior that can be proven or tested. Such properties become the
\emph{specification} against which enforcement can be validated; even enforcement mechanisms that do
not fulfill a property benefit from the ability to articulate \emph{why} they fail.

But most such properties are described at the level of a high-level programming language.
The stack as an abstraction exists below the level of
any language that uses it, and at the level of the ABI
\sna{Do we need to unpack the acronym ABI?}, the specification of a well-behaved
stack is minimal. It is not concerned with such questions as whether a caller's
frame should be readable or writable to its callee, leaving those concerns
to be determined by a high-level language built atop the stack.

Specifying stack safety in full generality would not merely be a matter of handling
a wide range of architectural variations. The basic security behavior that we expect
of the stack varies based on calling convention, compiler behavior, and the features of
whichever programming language originated the code.
(Contrast Azevedo de Amorim et al.'s \cite{DBLP:conf/post/AmorimHP18} work on heap safety,
which is broadly consistent across language semantics, and therefore
amenable to a very general approach.)

For instance, the current state-of-the-art stack-safety specification from Georges et al.
\cite{Georges22:TempsDesCerises} supposes a machine with only straightforward call-and-return
control flow, and therefore defines {\it Well-bracketed Control Flow} (WBCF) under the assumption that
a callee should always return to its caller. But, in the presence of tail-call elimination, a
return from a callee to a distant ancestor is perfectly reasonable.

In this work, we propose an explicit specification, built from first principles
in the hopes that it will be widely applicable, and demonstrate that it supports
an enforcement mechanism that does not yet have one.
We use the formal tools of language-based
security~\cite{sabelfeld2003language}, decomposing stack safety into
the {\em integrity} and {\em confidentiality} of the caller’s local state
and the callee's behavior during the callee's execution, plus the control-flow protection
of {\em well-bracketed control flow}~\cite{SkorstengaardSTKJFP} (\(\wbcf\)).
\ifexceptions
We also extend WBCF to a setting with exceptions,
in which there may be multiple points at which a callee can ``return.''
\fi
%
We treat stack safety as multiple properties, rather than one multi-faceted property, because
not all enforcement mechanisms attempt to implement all aspects of stack safety (see \cref{???}.)
%
This gives us a total of five core properties: \(\wbcf\),
caller integrity (\(\clri\)), caller confidentiality (\(\clrc\)),
callee integrity (\(\clei\)), and callee confidentiality (\(\clec\)), defined formally
in \cref{sec:props}.

We are not attempting to create a universal definition of stack safety for all systems.
Rather, we define the stack safety of a single, fairly realistic system, decomposed into
(1) a lightweight model of the system and its security-relevant features, from which we
derive the integrity, confidentiality, and control-flow requirements of each call and
return, (2) the application of that model to a particular target system, and (3)
the formal criteria for integrity, confidentiality, and control flow
to obey the requirements. This decomposition can be thought of as a step toward
a universal definition, in that the definitions in (3) capture fundamental
notions of security that can then be applied flexibly to other systems by varying
the model in (1) and (2).
%
\rb{FIXME: Contrast some universal definitions (or a parametric framework) with
  the concrete instance we study, again with its interest properly motivated.
  (Coming in key ideas?) No ``gold standard''?}
\apt{Yes, there are at least two different kinds of parameterization going on.
  One has to do with which properties we want to, or are able to, enforce; e.g.,
  for a language with tail calls or exceptions, we need a more complex definition of the WBCF property itself.
  Another has to do with how standard features, like argument passing, are implemented
  by particular hardware/ABIs -- which shouldn't show up in the property definition.
  We need to be clearer about this.}
\sna{In discussion above, I tried to carve out some separation there, but more may be needed.}

Our artifact contains Coq formalizations of our properties for illustrative purposes
as well as their testing implementation. There are no proofs---we use primarily Coq for the
QuickChick library and to ensure that the formalizations are unambiguous.

\subsection{Related Work: The Stack Safety Menagerie}

\apt{This section was a nice summary of enforcement approaches, but I am not
  yet seeing why it justifies so much paper real estate, especially if we only 
  actually apply our property definitions to (lazy) micropolicies. And in any case,
  it makes for a huge delay before getting to our work and contributions.}
\sna{I've tried out moving it here. To my mind, its main purpose is to justify
  our claims about the wide variation in stack-safety enforcement, and therefore
  to help argue that our properties not necessarily being universal is
  not a flaw.}

@@

\paragraph{Summary}

@@
